---
title: "Applying Partial Least Squares Regression (PLSR) To Explore Associations Between Cortical Thickness and Subvolume Using RANN Data"
author: "Melissa Nunez"
date: "10/15/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load RANN Data, echo=FALSE, include = FALSE}
# RQ: Is there an association between Cortical Thickness and Subcortical Volume? i.e can Cortical thickness predict cortical volume ?

#install.packages("caret")
library(caret)
library(pls)
library(ggplot2)
library(readxl)
library(knitr)
library(kableExtra)
library(gplots)

load("/Users/user/Dropbox/Melissa/PLSR/RANN_Data/data_creation_faMNI_sub2_fsaverages_volume_n173_20180817.Rdata")

```



```{r Step 3: Find optimal ncomps for each PLSR, echo=FALSE, include = FALSE}

# Bootstrap data 100 times
resamples <- lapply(1:100, function(i) sample(nrow(fs.thickness.resid), replace = T))

# Function takes in predictors, response, and indices of data (nums) for subsetting the predictors and response, then finds optimal number of components for plsr for each set of data, and returns coefficients of the plsr using the optimal number of components.
plsr_opt_coef <- function(x,y,nums)
  {

j=1
l <- list()

for (i in 1:length(nums)) {

x_boot <- x[nums[[i]],] 
y_boot <- y[nums[[i]],]
  
org_mod <- plsr(y_boot~x_boot, ncomp = 10, validation = "CV", segments = 5)

numcomps <- data.frame(sapply(1:10, function(x)(rowSums(data.frame(MSEP(org_mod, ncomp = x, intercept = FALSE)$val))/length(colnames(y)))[1]))
numcomps <- (cbind(c(1:10),numcomps))
names(numcomps) <- c("ncomp","Avg_MSEP")
min <- min(numcomps[,2])
opt_ncomps <- numcomps[numcomps$Avg_MSEP == min,][[1]]


coef_model <- coef(org_mod, ncomp = opt_ncomps)

l[[i]] <- coef_model
    j <- j + 1
    
}

list_opt_coefs <- l
assign("list_opt_coefs",list_opt_coefs,.GlobalEnv)

}

# Run plsr function
opt_coef_model <- plsr_opt_coef(fs.thickness.resid, fs.subvol.resid, resamples)


# Calculate mean and standard deviation of each coefficient
vec <- unlist(list_opt_coefs, use.names = FALSE)
DIM <- dim(list_opt_coefs[[1]])
n <- length(list_opt_coefs)

list.mean <- tapply(vec, rep(1:prod(DIM),times = n), mean)
attr(list.mean, "dim") <- DIM
list.mean <- as.data.frame(list.mean)

list.sd <- tapply(vec, rep(1:prod(DIM),times = n), sd)
attr(list.sd, "dim") <- DIM
list.sd <- as.data.frame(list.sd)

# Calculate z-score for each coeefficient
z.score_mat <- list.mean/list.sd

colnames(z.score_mat) <- colnames(fs.subvol.resid)
rownames(z.score_mat) <- colnames(fs.thickness.resid)

# Create matrix of significant associations
thresh=as.matrix(z.score_mat)
thresh[abs(thresh)<1.96]<-0
thresh <- thresh[as.logical(rowSums(thresh != 0)), as.logical(colSums(thresh != 0))]
```


#### Descriptive Statistics: Gender and Education
```{r report, echo = FALSE, warning=FALSE, message = FALSE}
# Tables for descriptive stats
Count.Gender <- addmargins(table(fsdat$Gender))
Count.Gender <- data.frame(c("Female","Male","Total"), c(99,74,173))
colnames(Count.Gender) <- c("Gender","Count")
Count.Gender %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")

Edu.Summary <- data.frame(as.matrix(round(summary(fsdat$Education),2)))
colnames(Edu.Summary) <- c("Years of Education")
Edu.Summary %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")
```


#### Age Summary
```{r statreport, echo = FALSE, echo = FALSE, warning=FALSE, message = FALSE}
Age.Summary <- data.frame(as.matrix(round(summary(fsdat$Age),2)))
colnames(Age.Summary) <- c("Baseline Age")
Age.Summary %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")
```

RANN data was used to explore associations between cortical thickness and subvolume measures. One hundred boostraped sets were generated.
On each dataset, partial least squares regression was applied with subvolume measures as the outcomes and cortical thickness measures as 
the predictors. 5-fold cross validation was used to determine the optimal number of components that should be used for each set of
outcomes. Beta coefficients were averaged across all 100 datasets and standardized. The following heat map summarizes the significant
associations.

```{r heatmap, fig.height=8, fig.width=10, fig.align="center", message = FALSE, echo = FALSE, warning=FALSE,}
heatmap.2(thresh, scale="none",trace='none',col=bluered(200),
          breaks=c(-100:100)/100*5, margins = c(15,22), dendrogram = "none")
```
