---
title: "Applying Partial Least Squares Regression (PLSR) To Explore Associations Between Cortical Thickness and Subvolume Using ADNI Data"
author: "Melissa Nunez"
date: "10/14/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Data Preparation, include = FALSE}
library(caret)
library(pls)
library(ggplot2)
library(readxl)
library(knitr)
library(kableExtra)
library(gplots)

ADNI_Set1 <- read_excel("/Users/user/Desktop/WorkDirectory/imputation/ADNI/dataset/ADNI\ 6\ datasets\ creation\ 2018_06_04.xlsx", sheet = 2)

# Subset
ADNI_Set1 <- ADNI_Set1[which(ADNI_Set1$dataset1 == 1 & ADNI_Set1$month_dataset1 == 0),]
head(ADNI_Set1,10)

# Group variables
CT <- grep("CT",names(ADNI_Set1))
SV <- grep("SV_",names(ADNI_Set1))
SV <- SV[17:66]

names(ADNI_Set1)[343]
length(SV)

# Some more subsetting
CT_Set <- data.frame(ADNI_Set1[,CT])
SV_Set <- data.frame(ADNI_Set1[,SV])
head(CT_Set, 10)

```


```{r Bootstrap, include = FALSE, echo= FALSE}
# Bootstrap the data 100 times. Resamples is a list with 100 elements, each element has the indices of one boostrap set
set.seed(001)
resamples <- lapply(1:100, function(i) sample(nrow(ADNI_Set1), replace = T))
typeof(as.matrix(SV_Set[resamples[[1]],]))

```



```{r Analysis, echo=FALSE, include = FALSE}
# Create function that runs plsr for each boostrap dataset, and gives the beta coefficients  for each regression based on the optimal number of components
plsr_opt_coef <- function(x,y,nums)
  {

j=1
l <- list()

for (i in 1:length(nums)) {

x_boot <- as.matrix(x[nums[[i]],])
y_boot <- as.matrix(y[nums[[i]],])
  
org_mod <- plsr(y_boot~x_boot, ncomp = 10, validation = "CV", segments = 5)

numcomps <- data.frame(sapply(1:10, function(x)(rowSums(data.frame(MSEP(org_mod, ncomp = x, intercept = FALSE)$val))/length(colnames(y)))[1]))
numcomps <- (cbind(c(1:10),numcomps))
names(numcomps) <- c("ncomp","Avg_MSEP")
min <- min(numcomps[,2])
opt_ncomps <- numcomps[numcomps$Avg_MSEP == min,][[1]]


coef_model <- coef(org_mod, ncomp = opt_ncomps)

l[[i]] <- coef_model
    j <- j + 1
    
}

list_opt_coefs <- l
assign("list_opt_coefs",list_opt_coefs,.GlobalEnv)

}


# Run plsr function
opt_coef_model <- plsr_opt_coef(CT_Set, SV_Set, resamples)


# Calculate mean and standard deviation for coefficients
vec <- unlist(list_opt_coefs, use.names = FALSE)
DIM <- dim(list_opt_coefs[[1]])
n <- length(list_opt_coefs)

list.mean <- tapply(vec, rep(1:prod(DIM),times = n), mean)
attr(list.mean, "dim") <- DIM
list.mean <- as.data.frame(list.mean)

list.sd <- tapply(vec, rep(1:prod(DIM),times = n), sd)
attr(list.sd, "dim") <- DIM
list.sd <- as.data.frame(list.sd)

# Calculate z-score for coefficients
z.score_mat <- list.mean/list.sd
dim(z.score_mat)

# Rename columns
colnames(z.score_mat) <- colnames(SV_Set)
rownames(z.score_mat) <- colnames(CT_Set)

# Create matrix of significant coefficients
thresh <- as.matrix(z.score_mat)
thresh[abs(thresh)<1.96] <- 0
thresh <- thresh[as.logical(rowSums(thresh != 0)), as.logical(colSums(thresh != 0))]
```



#### Descriptive Statistics: Gender and Education
```{r report, echo = FALSE, warning=FALSE, message = FALSE}
# Summary table of gender and Education
Count.Gender <- addmargins(table(ADNI_Set1$PTGENDER))
Count.Gender <- data.frame(c(1,2,"Total"), c(754,616,1370))
colnames(Count.Gender) <- c("Gender","Count")
Count.Gender %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")

Count.Edu <- addmargins(table(ADNI_Set1$EDU))
Count.Edu <- data.frame(c("High","Low","Total"), c(1168,202,1370))
colnames(Count.Edu) <- c("Education","Count")
Count.Edu %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")

```


#### Age Summary
```{r statreport, echo = FALSE, warning=FALSE, message = FALSE}
# Summary of Age at Baseline (Table)
Age.Summary <- data.frame(as.matrix(summary(ADNI_Set1$Age_at_baseline)))
colnames(Age.Summary) <- c("Baseline Age")
Age.Summary %>%
  kable() %>%
    kable_styling(bootstrap_options = "striped")
```

ADNI data was used to explore associations between cortical thickness and subvolume measures. One hundred boostraped sets were generated. On each dataset, partial least squares regression was applied with subvolume measures as the outcomes and cortical thickness measures as the predictors. 5-fold cross validation was used to determine the optimal number of components that should be used for each set of outcomes. Beta coefficients were averaged across all 100 datasets and standardized. The following heat map summarizes the significant associations.

```{r heatmap, fig.height=8, fig.width=13, fig.align="center", echo = FALSE, warning=FALSE, message = FALSE}
heatmap.2(thresh, scale="none",trace='none',col=bluered(200),
          breaks=c(-100:100)/100*5, margins = c(14,14), dendrogram = "none", xlab = "Subvolume", ylab = "Cortical Thickness", key.par=list(mar=c(6,6,6,6)))
```


